Can Auxiliary Indicators Improve COVID-19 Forecasting and Hotspot Prediction?

### 2021-10-04 forecast/paper/response-to-reviewers.tex

\begin{resp}
  We have tried to focus mainly on simple time series models in this paper
  rather than comparing to other more elaborate models. A careful
  (out-of-sample) comparison of forecasting performance for state-of-the-art
  models can be found in reference 6 (Cramer et al.). That paper focuses on
  forecasting deaths rather than cases and at the state-level rather than
  disaggregated HRRs. However, their ranking (Figure 2 of Cramer et al.) puts the
  ``COVIDhub-Baseline'' about 10th out of 27 models. As we discuss in the
  manuscript, that baseline is  
  the
  same as the baseline we use for forecasting in this paper: it propagates today
  forward and resamples the residuals to create quantiles (we have expanded the
  description on page 6 to be more explicit). All our ``simplistic'' models beat
  this baseline (the AR(3) does as well). For comparison, IHME (along with 17 other
  ``state-of-the-art'' models) does not. We would argue that using any of the
  models in this paper would be preferable to relying on many of the
  ``widely-used'' models if the goal is forecast accuracy.

  $\bullet$ We have added a sentence to the introduction (around line 40) that
  describes the performance of simple models relative to widely-used models as
  shown in Cramer et al.

  $\bullet$ We have expanded our description of our baseline model (around line
  457). In particular, we emphasize that it is the same as that used by the
  COVID-19 Forecast Hub.

  $\bullet$ Here are adding two figures
    that compare our forecasts with state-of-the-art forecasts
  submitted to the COVID-19 Forecast Hub. The first overlays the
  scores from the submissions on top of Figure 3 from the manuscript. We should
  note that the submitted forecasts are at the state level, while ours are at
  the HRR level, but once scaled by the baseline, should be roughly comparable.
  Only those forecasters that submitted at least 300 forecasts (roughly 50
  states over 6 weeks during the 6 months examined) are shown. From this Figure
  it is clear that all of the forecasters considered in this paper beat most
  submissions. In the second figure, we use the Geometric Mean scaled by the
  baseline, which should be a more meaningful comparison, as argued elsewhere in
  the supplement. By this metric, the methods employed in this
  paper would rank in the top 2 of all submitted forecasts, beating the
  COVIDhub-ensemble at all forecast 
  horizons. The improvement of the best indicator assisted model over the AR model
  is roughly the same as the improvement of the AR model over the
  COVIDhub-ensemble.

  \begin{figure}[tb]
    \centering
    \includegraphics[width=.9\textwidth]{fig/compare-to-hub-mean-1.pdf}
    \caption{This figure reproduces Figure 3 in the main paper but overlays
      scores for the forecasts submitted to the COVID-19 Forecast Hub. Grey
      lines correspond to the various teams that submitted during period our
      evaluation period. We have highlighted the COVIDhub-ensemble, which is the
      official forecast of the CDC.}
    \label{fig:compare-to-hub-mean}
  \end{figure}

  \begin{figure}[tb]
    \centering
    \includegraphics[width=.9\textwidth]{fig/compare-to-hub-geomean-1.pdf}
    \caption{This figure is more like Figure S6 in the Supplement. In this case,
      we add 1 to both the forecaster WIS and the baseline WIS before scaling
      (to allow forecasters that achieve 0 error to appear), and we overlay
      scores for the forecasts submitted to the COVID-19 Forecast Hub. Grey
      lines correspond to the various teams that submitted during period our
      evaluation period. We have highlighted the COVIDhub-ensemble, which is the
      official forecast of the CDC.}
    \label{fig:compare-to-hub-geomean}
  \end{figure}
\end{resp}

### 2021-10-04 forecast/paper/paper.tex
clean up figure 5, move hub comparisons out

% Please add a significance statement to explain the relevance of your work
\significancestatement{Forecasting is a critical element in the public health
  response to any fast-moving epidemic or pandemic.  Predicting the future
  spread of some disease process, based on disease activity in the recent
  past---say, using autoregressive modeling from the time series literature---is 
  a basic and widely-used method.  In this paper we study whether including
  auxiliary indicators as additional features in such an autoregressive model
  can be useful in COVID-19 case forecasting and hotspot prediction.  We find
  that indicators based on de-identified medical insurance claims, self-reported
  symptoms through online surveys, and COVID-related internet search activity
  all provide a nontrivial improvement in forecast accuracy, though their
  utility depends on the pandemic's local dynamics (upswing, downswing, or flat
  period) at prediction time. 
}

@@

While forecasting carries a rich literature that offers a wide range of
techniques, see e.g., \cite{Hyndman:2018}, we purposely constrain ourselves to
very simple models, avoiding common enhancements such as order selection,
correction of outliers/anomalies in the data, and inclusion of regularization or
nonlinearities.
% Addition in response to Reviewer 1
For the same reason, we do not account for any environmental factors that may
well aid in forecasting---age structure, variants, holidays,
reporting patterns, vaccination, public health measures, etc.
% End addition
That said, analyses of forecasts submitted to the COVID-19
Forecast Hub \cite{ForecastHub} by a large community of modelers have shown that
simple, robust models have consistently been among the best-performing over the
pandemic, including time series models similar to those we
consider in what follows.
% Added in response to Reviewer 2
\cite{Cramer:2021} investigates the performance of submitted forecasts for
state-level deaths and shows that time series models perform among the top 5.
In the Supplement, we compare the time series models investigated in this paper
to the submitted case forecasts over the same period. The comparison is not
particularly flattering: the simple models examined here outperform most
submitted models and approach the accuracy of the COVIDHub Ensemble used by the
CDC.

@@

Figure~\ref{fig:forecast} (left panel) displays evaluation results for
forecasting, stratified by ahead value and averaged over all HRRs and forecast
dates.  Shown is the average WIS for each forecast model divided by that from a
baseline model. The baseline model is a flat-line forecaster which forms its
median forecast by using the most recent value $Y_{\ell,t}$ for all aheads
$Y_{\ell,t+a}$, with predicted quantiles defined by the empirical distribution
of the residuals from this median forecast over recent history. This is the same
baseline model as in the COVID-19 Forecast Hub. Here, we use the baseline model
in order to scale mean WIS to put it on an interpretable, unitless
scale.  In the figure, we can see that all curves are below 1, which means
(smaller WIS is better) that all of the models, including AR, outperform the
baseline on average over the forecasting period.  On the other hand, the models
deliver at best an improvement of about 20\% in average WIS over the baseline
model, with this gap narrowing to about 10\% at the largest ahead values,
illustrating the difficulty of the forecasting problem.

>>>>

% Please add a significance statement to explain the relevance of your work
\significancestatement{Forecasting is a critical element in the public health
  response to any fast-moving epidemic or pandemic.  Predicting the future
  spread of some disease process, based on disease activity in the recent
  past---say, using what is known as autoregressive modeling in the time series
  literature---is a basic and widely-used method.  In this paper we study
  whether including auxiliary indicators as additional features in such an
  autoregressive model can be useful in COVID-19 case forecasting and hotspot
  prediction.  We find that indicators based on de-identified medical insurance 
  claims, self-reported symptoms through online surveys, and COVID-related
  internet search activity
  % all provide a nontrivial boost in accuracy,
  all provide a nontrivial improvement in forecast accuracy,
  though
  their utility generally depends on the pandemic's local dynamics (upswing,
  downswing, or flat period) at prediction time.
}

@@

While forecasting carries a rich literature that offers a wide range of
techniques, see e.g., \cite{Hyndman:2018}, we purposely constrain ourselves to
very simple models, avoiding common enhancements such as order selection,
correction of outliers/anomalies in the data, and inclusion of regularization or
nonlinearities.
% Addition in response to Reviewer 1
For the same reason, we do not account for any environmental factors that may
well aid in forecasting---age structure, variants, holidays,
reporting patterns, vaccination, public health measures, etc.
% End addition
That said, analyses of forecasts submitted to the COVID-19
Forecast Hub \cite{ForecastHub} by a large community of modelers have shown that
simple, robust models have consistently been among the best-performing over the
pandemic, including time series models similar to those we
consider in what follows.
% Added in response to Reviewer 2
\cite{Cramer:2021} investigates the performance of submitted forecasts for
state-level deaths and shows that time series models perform among the top 5.
Our internal evaluations suggest that the same is true (at least, if not to a
greater extent) for case forecasting.

@@

Figure~\ref{fig:forecast} (left panel) displays evaluation results for
forecasting, stratified by ahead value and averaged over all HRRs and forecast
dates.  Shown is the average WIS for each forecast model divided by that from a
% change since review
baseline model.
The baseline model is a flat-line forecaster which forms its median forecast by
using the most recent value $Y_{\ell,t}$ as the median forecast for all aheads
$Y_{\ell,t+a}$ . To create
the quantiles, we use the empirical quantiles of the residuals from this
forecast over the available history. 
% old version
% baseline model,
% which is basically a flat-line forecaster (its median forecast
% for $Y_{\ell,t+a}$ is always $Y_{\ell,t}$, with predicted quantiles defined
% around this based on historical variation).
This is the same baseline model as
in the COVID-19 Forecast Hub. Here, we use the baseline model in order to scale
mean WIS so that it gets put on an interpretable, unitless scale.  In the
figure, we can see that all curves are below 1, which means (smaller WIS is
better) that all of the models, including AR, outperform the baseline on average
over the forecasting period.  On the other hand, the models deliver at best an
improvement of about 20\% in average WIS over the baseline model, with this gap
narrowing to about 10\% at the largest ahead values, illustrating the difficulty 
of the forecasting problem.

### 2021-10-04 forecast/paper/supplement.Rmd
clean up figure 5, move hub comparisons out

```{r upswing-histogram, include=FALSE, fig.width=9, fig.height=4.5}
# Figure 4
up_down_df %>% 
  filter(period != "jm") %>%
  group_by(forecaster, udf) %>%
  ggplot(aes(AR_wis - wis, fill = forecaster)) +
  geom_histogram(bins = 100) +
  facet_grid(udf ~ forecaster) +
  theme_bw() +
  ylab("Count") +
  theme(legend.position = "none") +
  scale_fill_manual(values = fcast_colors) +
  scale_y_log10(breaks = c(1,100,10000), 
                labels = trans_format("log10", math_format(10^.x))) +
  coord_cartesian(ylim=c(1,2e4)) +
  xlab("AR WIS - forecaster WIS") +
  geom_vline(xintercept = 0)
```

@@

```{r hotspots-upswing-downswing-remake, fig.cap="REMAKE: Classification and loglikelihood separated into periods of upswing, downswing, and flat cases. The indicator assisted models always have lower classification error relative to the null classifier, while the AR actually does worse in down and flat periods. In terms of loglikelihood, all forecasters have lower likelihood than the null classifier during up periods. In down and flat periods, the indicators generally improve over the AR, while the are worse during up periods."}
hot_udf <- inner_join(
  hotspots_honest %>% filter(period != "jm"),
  up_down %>% select(geo_value, target_end_date, udf))

@@

con_tab <- hot_udf %>% 
  filter(!is.na(actual)) %>%
  mutate(pred = value > cutoff) %>%
  group_by(forecaster, udf) %>%
  summarise(m = mean(pred != actual) - .5) %>%
  ungroup()

@@

llike_tab <- hot_udf %>%
  filter(!is.na(actual)) %>%
  mutate( # kill prob 0-1 predictions
    value = case_when(
      value < 1e-8 ~ 1e-8,
      value > 1-1e-8 ~ 1-1e-8,
      TRUE ~ value
    ),
    llike = (actual == 1) * log(value) + (actual == 0) * log(1 - value),
    nulldev = (actual == 1) * log(.5) + (actual == 0) * log(1 - .5)) %>%
  group_by(forecaster, udf) %>%
  summarise(m = mean(llike) - mean(nulldev)) %>%
  ungroup()

@@

bind_rows(con_tab %>% mutate(err = "Classification error"), 
          llike_tab %>% mutate(err = "Log likelihood")) %>%
  ggplot(aes(udf, m , fill = forecaster)) +
  geom_bar(width = 0.6, position = position_dodge(width=0.6),
           stat = "identity") +
  scale_fill_manual(values = fcast_colors, guide = guide_legend(nrow = 1)) +
  #scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_bw() +
  geom_hline(yintercept = 0) +
  facet_wrap(~err, scales = "free_y") +
  ylab("Improvement relative to the null classifier") +
  xlab("") +
  theme(legend.position = "bottom", legend.title = element_blank())
```

@@

```{r leading-only, include=FALSE, fig.width=6, fig.height=4}
# Figure 6 in the manuscript
df2 %>%
  filter(period != "jm") %>%
  group_by(forecaster, udf) %>%
  summarise(
    leadingness = cor(AR_wis - wis, leading, use = "pairwise.complete.obs"),
    laggingness = cor(AR_wis - wis, lagging, use = "pairwise.complete.obs")) %>%
  ggplot(aes(udf, leadingness, fill = forecaster)) +
  geom_bar(width = 0.6, position = position_dodge(width=0.6),
           stat = "identity") +
  geom_hline(yintercept = 0) +
  scale_fill_manual(values = fcast_colors) +
  guides(fill = guide_legend(nrow = 1)) +
  theme_bw() +
  ylab("Correlation") +
  xlab("") +
  theme(legend.position = c(1,-.15), 
        legend.justification = "right",
        legend.title = element_blank(), 
        legend.text = element_text(size = 10),
        plot.margin = margin(b = 20, l = 4),
        legend.background = element_rect(fill = "transparent"))
```

@@

```{r compare-to-hub-mean, fig.height = 8.5, fig.cap="This figure reproduces Figure 3 in the main paper but overlays scores for the forecasts submitted to the COVID-19 Forecast Hub. Grey lines correspond to the various teams that submitted during period our evaluation period. We have highlighted the COVIDhub-ensemble, which is the official forecast of the CDC."}
# Same as Fig 3 in the paper
df <- summarizer(fcasts_honest %>% filter(period != "jm") %>%
                   group_by(ahead, forecaster), "wis", NULL, 
                 "strawman_wis", Mean, c("aggr","scale"))
ggplot(df) +
  geom_line(aes(ahead, wis, color = forecaster)) + 
  geom_point(aes(ahead, wis, color = forecaster)) +
  theme_bw() +
  geom_hline(yintercept = 1, size = 1.5) +
  xlab("Days ahead") +
  ylab("Mean WIS (relative to baseline)") +
  scale_color_manual(values = c(fcast_colors, "CH-ensemble" = "lightblue"),
                     guide = guide_legend(nrow = 2)) +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  geom_line(data = case_scores, 
            aes(ahead * 7, rel_wis, group = forecaster),
            color = "grey70") +
  geom_line(data = case_scores %>% filter(forecaster == "COVIDhub-ensemble"),
            aes(ahead * 7, rel_wis), color = "lightblue", size = 1.5) +
  scale_y_log10()
```

@@

```{r compare-to-hub-geomean, fig.height = 8.5, fig.cap="This figure is similar to Figure \\ref{fig:fcast-adjusted}. In this case, we add 1 to both the forecaster WIS and the baseline WIS before scaling (to allow forecasters that achieve 0 error to appear), and we overlay scores for the forecasts submitted to the COVID-19 Forecast Hub. Grey lines correspond to the various teams that submitted during period our evaluation period. We have highlighted the COVIDhub-ensemble, which is the official forecast of the CDC."}
# Like supplement GeoMean figure, but adding 1
df2 <- fcasts_honest %>% 
  filter(period != "jm") %>%
  group_by(ahead, forecaster) %>%
  summarise(wis = GeoMean((wis + 1) / (strawman_wis + 1)))

>>>>

```{r upswing-histogram, include=FALSE, fig.width=9, fig.height=4.5}
# Fgure 4
up_down_df %>% 
  filter(period != "jm") %>%
  group_by(forecaster, udf) %>%
  ggplot(aes(AR_wis - wis, fill = forecaster)) +
  geom_histogram(bins = 100) +
  facet_grid(udf ~ forecaster) +
  theme_bw() +
  ylab("Count") +
  theme(legend.position = "none") +
  scale_fill_manual(values = fcast_colors) +
  scale_y_log10(breaks = c(1,100,10000), 
                labels = trans_format("log10", math_format(10^.x))) +
  coord_cartesian(ylim=c(1,2e4)) +
  xlab("AR WIS - forecaster WIS") +
  geom_vline(xintercept = 0)
```

@@

```{r hotspots-upswing-downswing-remake, fig.cap="REMAKE: Classification and loglikelihood separated into periods of upswing, downswing, and flat cases. The indicator assisted models always have lower classification error relative to the null classifier, while the AR actually does worse in down and flat periods. In terms of loglikelihood, all forecasters have lower likelihood than the null classifier during up periods. In down and flat periods, the indicators generally improve over the AR, while the are worse during up periods."}
# mycummean <- function(x) {
#   isgood <- ! is.na(x)
#   denom <- cumsum(isgood)
#   x[!isgood] <- 0
#   num <- cumsum(x)
#   y <- num / denom
#   y[is.na(y)] <- .5 # deal with empties at the beginning
#   y
# }
# 
# null_classifier <- readRDS(here("data", 
#                                 "all_signals_wide_as_of_2020-05-18.RDS")) %>%
#   rename(jhu = `value+0:jhu-csse_confirmed_7dav_incidence_prop`) %>%
#   select(geo_value, time_value, jhu) %>%
#   group_by(geo_value) %>%
#   arrange(time_value) %>%
#   mutate(lag_value = lag(jhu, 7),
#          apc = (jhu - lag_value) / lag_value,
#          hot = case_when(
#            apc > .25 & lag_value > 30 ~ 1,
#            apc <= .25 & lag_value > 30 ~ 0),
#          null_class = mycummean(hot))

hot_udf <- inner_join(
  hotspots_honest %>% filter(period != "jm"),
  up_down %>% select(geo_value, target_end_date, udf))

@@

con_tab <- hot_udf %>% 
  filter(!is.na(actual)) %>%
  mutate(pred = value > cutoff) %>%
  group_by(forecaster, udf) %>%
  summarise(m = (mean(pred != actual) - cutoff) / cutoff,
            err = mean(pred != actual)) %>%
  ungroup()

@@

llike_tab <- hot_udf %>%
  filter(!is.na(actual)) %>%
  # left_join(null_classifier %>% select(geo_value, time_value, null_class) %>%
  #             rename(target_end_date = time_value)) %>%
  mutate( # kill prob 0-1 predictions
    value = case_when(
      value < 1e-8 ~ 1e-8,
      value > 1-1e-8 ~ 1-1e-8,
      TRUE ~ value),
    # null_class = case_when(
    #   null_class < 1e-8 ~ 1e-8,
    #   null_class > 1-1e-8 ~ 1-1e-8,
    #   TRUE ~ null_class),
    llike = (actual == 1) * log(value) + (actual == 0) * log(1 - value),
    nulldev = (actual == 1) * log(cutoff) + 
      (actual == 0) * log(1 - cutoff)) %>%
  group_by(forecaster, udf) %>%
  summarise(m = (mean(llike) - mean(nulldev)) / abs(mean(nulldev)),
            mm = mean(llike),
            mmm = mean(nulldev)) %>%
  ungroup()

@@

bind_rows(con_tab %>% mutate(err = "Classification error"), 
          llike_tab %>% mutate(err = "Log likelihood")) %>%
  ggplot(aes(udf, m , fill = forecaster)) +
  geom_bar(width = 0.6, position = position_dodge(width=0.6),
           stat = "identity") +
  scale_fill_manual(values = fcast_colors, guide = guide_legend(nrow = 1)) +
  #scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_bw() +
  geom_hline(yintercept = 0) +
  scale_y_continuous(labels = scales::percent_format()) +
  facet_wrap(~err, scales = "free_y") +
  ylab("Improvement relative to the null classifier") +
  xlab("") +
  theme(legend.position = "bottom", legend.title = element_blank())
```

@@

```{r leading-only, include=FALSE, fig.width=6, fig.height=4}
# Figure 5 in the manuscript
df2 %>%
  filter(period != "jm") %>%
  group_by(forecaster, udf) %>%
  summarise(
    leadingness = cor(AR_wis - wis, leading, use = "pairwise.complete.obs"),
    laggingness = cor(AR_wis - wis, lagging, use = "pairwise.complete.obs")) %>%
  ggplot(aes(udf, leadingness, fill = forecaster)) +
  geom_bar(width = 0.6, position = position_dodge(width=0.6),
           stat = "identity") +
  geom_hline(yintercept = 0) +
  scale_fill_manual(values = fcast_colors[1:5]) +
  guides(fill = guide_legend(nrow = 1)) +
  theme_bw() +
  ylab("Correlation") +
  xlab("") +
  theme(legend.position = c(1,-.15), 
        legend.justification = "right",
        legend.title = element_blank(), 
        legend.text = element_text(size = 10),
        plot.margin = margin(b = 20, l = 4),
        legend.background = element_rect(fill = "transparent"))
```

@@

```{r compare-to-hub-mean, fig.height = 8.5, fig.cap="This figure reproduces Figure 3 in the main paper but overlays scores for the forecasts submitted to the COVID-19 Forecast Hub. Grey lines correspond to the various teams that submitted during period our evaluation period. We have highlighted the COVIDhub-ensemble, which is the official forecast of the CDC.", include=FALSE}
# Same as Fig 3 in the paper
df <- summarizer(fcasts_honest %>% filter(period != "jm") %>%
                   group_by(ahead, forecaster), "wis", NULL, 
                 "strawman_wis", Mean, c("aggr","scale"))
ggplot(df) +
  geom_line(aes(ahead, wis, color = forecaster)) + 
  geom_point(aes(ahead, wis, color = forecaster)) +
  theme_bw() +
  geom_hline(yintercept = 1, size = 1.5) +
  xlab("Days ahead") +
  ylab("Mean WIS (relative to baseline)") +
  scale_color_manual(values = c(fcast_colors, "CH-ensemble" = "lightblue"),
                     guide = guide_legend(nrow = 2)) +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  geom_line(data = case_scores, 
            aes(ahead * 7, rel_wis, group = forecaster),
            color = "grey70") +
  geom_line(data = case_scores %>% filter(forecaster == "COVIDhub-ensemble"),
            aes(ahead * 7, rel_wis), color = "lightblue", size = 1.5) +
  scale_y_log10()
```

@@

```{r compare-to-hub-geomean, fig.height = 8.5, fig.cap="This figure is similar to Figure \\ref{fig:fcast-adjusted}. In this case, we add 1 to both the forecaster WIS and the baseline WIS before scaling (to allow forecasters that achieve 0 error to appear), and we overlay scores for the forecasts submitted to the COVID-19 Forecast Hub. Grey lines correspond to the various teams that submitted during period our evaluation period. We have highlighted the COVIDhub-ensemble, which is the official forecast of the CDC.", include=FALSE}
# Like supplement GeoMean figure, but adding 1
df2 <- fcasts_honest %>% 
  filter(period != "jm") %>%
  group_by(ahead, forecaster) %>%
  summarise(wis = GeoMean((wis + 1) / (strawman_wis + 1)))
